---
title: "The package countries is now on CRAN"
description: |
  After a year of development, I am happy to annouce that my package can now be downloaded directly from R's official repository. And it has some cool new functionalities!
author:
  - name: Francesco S. Bellelli
date: 2023-09-24
categories: 
  - coding
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    self_contained: false
preview: R_logo.png
---

```{r, include = FALSE}
library(countries)
```



In practical terms, it means that you can now download and load the package `countries` simply with the commands:


```{r, eval = FALSE}
install.packages("countries")
library(countries)
```

This official release also coincides with some major additions to the package. For example, the function `country_info` allows to download information from the [countries REST API](https://restcountries.com/) and the function `quick_map` makes it super easy to produce [country chloropleth maps](https://en.wikipedia.org/wiki/Choropleth_map) thanks to fuzzy matching of country names. Here, have a look at how easy is to get information on which side of the road people drive in every country, and plot it on a map:


```{r, dpi = 700, warning = FALSE, message = FALSE, fig.asp=0.5}
# downloading information on driving side
example_data <- country_info(fields = c("car"))

# make a map
quick_map(example_data, plot_col = "car.side")
```


For more information on the package and the new functionalities, consult the [website of the package](https://fbellelli.github.io/countries/).